```{r}
library(nflverse)
library(tidyverse)
library(ggplot2)
library(cfbfastR)
library(fuzzyjoin)
library(rvest)
library(randomForest)
library(Metrics)
library(tidytext)
library(gt)
library(gtExtras)
library(grid)

```



```{r}
#Loading and filtering draft data
load_filtered_draft <- function(season) {
  load_draft_picks(seasons = season) %>%
    filter(position != '') %>%
    mutate(season = season)
}

past_years <- 2015:2024
past_drafts <- map_df(past_years, load_filtered_draft)

```


```{r}
#Collecting the NFL stats of all drafted RBs

draft <- load_draft_picks()

drafted_rbs <- draft %>%
  filter(position == "RB", season >= 2015) %>% mutate(
    cfb_player_id = if_else(
      is.na(cfb_player_id),
      str_replace_all(str_to_lower(pfr_player_name), " ", "-"),
      cfb_player_id
    ))


stats <- load_player_stats(seasons = c(2016:2024), stat_type = "offense") %>% filter(position == "RB", season_type == "REG") %>% select(-c(recent_team,week,opponent_team,season_type)) %>% group_by(player_id, player_name, player_display_name, position_group, headshot_url, season) %>% 
  summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop")

rb_first_3 <- stats %>%
  group_by(player_id) %>%
  arrange(season) %>%
  slice_head(n = 3) %>%
  mutate(season = row_number()) %>%
  ungroup() %>% rename(gsis_id = player_id)




rb_wide <- rb_first_3 %>%
  select(gsis_id:headshot_url, where(is.numeric)) %>%
  pivot_wider(
    names_from = season,
    values_from = where(is.numeric),
    names_sep = "_"
  )

drafted_rbs_with_stats <- drafted_rbs %>%
  left_join(rb_wide, by = c("gsis_id")) %>% select(-c(player_name:headshot_url)) %>% 
  group_by(season, round, pick, team, gsis_id, pfr_player_id, pfr_player_name, college) %>% 
  summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop") %>% 
  select(-c(pass_completions:pass_ints, to, allpro:dr_av, def_solo_tackles:def_sacks, attempts_1:dakota_3, completions_1:completions_3)) %>% 
  rename_with(.fn = ~ paste0("college_", .), .cols = 9:16)


```






```{r}
#looking at SRS for teams to assign relative team values for players
#This would help assess in the model if the player's success is related to their team's success (SRS Score)
#I.E. is a Alabama or Georgia RB actually good? or are they just on a stacked team?

srs_all_years <- map_dfr(2014:2025, ~ cfbd_ratings_srs(year = .x))%>%
  mutate( team = str_replace(team, "San José State", "San Jose State"),
          team = str_replace(team, "\\bState\\b", "St."), 
          team = if_else(team == "Miami", "Miami (FL)", team),
          year = year + 1)



#Manually had to fill in missing data


drafted_rbs_with_stats_srs <- drafted_rbs_with_stats %>%
  left_join(srs_all_years, by = c("college" = "team", "season" = "year")) %>% 
  select(-c(division, ranking)) %>% mutate(
    college = ifelse(college == "Boston Col.", "Boston College", college)
  ) %>% 
  relocate("rating", .before = "college_age") %>% mutate(
    rating = case_when(
      college == "Texas-El Paso" & season == 2018 ~ -11.4,     # UTEP
      college == "Mississippi" & season == 2019 ~ 1.0,         # Ole Miss
      college == "Mississippi" & season == 2024 ~ 14.3,        # Ole Miss
      college == "Mississippi" & season == 2023 ~ 12.5,        # Ole Miss
      college == "Appalachian St." & season == 2020 ~ 9.5,     # App State
      college == "North Carolina St." & season == 2018 ~ 10.1, # NC St.
      college == "North Carolina St." & season == 2019 ~ 6.8,  # NC St.
      college == "Ala-Birmingham" & season == 2024 ~ -14.3,    #UAB
      college == "Miami (FL)" & season == 2021 ~ 6.56, #Miami
      college == "Boston College" & season == 2021 ~ -0.04,
      college == "Florida St." & season == 2021 ~ -6.6,
      college == "Memphis" & season == 2021 ~ -1.29,
      college == "Navy" & season == 2021 ~ -10.08,
      college == "New Mexico St." & season == 2021 ~ -27.68,
      college == "Arizona St." & season == 2021 ~ 1.75,
      college == "UCLA" & season == 2021 ~ -0.11,
      college == "Utah" & season == 2021 ~ 0.67,
      college == "Appalachian St." & season == 2021 ~ 5.47,
      college == "Central Florida" & season == 2025 ~ 6.2,
      college == "Clemson" & season == 2021 ~ 16.82,
      college == "North Carolina" & season == 2021 ~ 7.03,
      college == "Arizona" & season == 2021 ~ -14.08,
      college == "Virginia Tech" & season == 2021 ~ 1.51,
      college == "Cinncinati" & season == 2021 ~ 14.92,
      
      TRUE ~ rating
    )
  ) %>% 
  select(season:team, pfr_player_name:rating, rushing_yards_1:rushing_yards_3, receiving_yards_1:receiving_yards_3) %>% 
  mutate(missing_rating = is.na(rating),
         rating = rating) %>% 
  relocate(missing_rating, .after = rating)


```






#Gathering PFF Stats and Base Stats for college

```{r}

#Data Accessed from PFF, using a paid-PFF Plus account - Private data, and is illegal to redistribute on GitHub

folder_path <- "rushing sums"

# List all CSVs in folder
csv_files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)

# Load each file, add season from filename
rushing_data <- csv_files %>%
  lapply(function(file) {
    df <- read_csv(file)
    
    # Extract last two digits and convert to year
    season_year <- str_extract(file, "\\d{2}(?=\\.csv$)")
    season <- as.numeric(season_year) + 2000
    
    df <- df %>% mutate(season = season)
    return(df)
  }) %>%
  bind_rows()



rb_names_pff <- drafted_rbs_with_stats_srs %>%
  select(season, pfr_player_name) %>%
  distinct()

rb_names_realpff <- rushing_data %>% 
  select(season, player) %>% 
  distinct()


rb_names_realpff_clean <- rb_names_realpff %>%
  filter(player != "Drake Johnson")

# Step 1: Fuzzy join on player name only
fuzzy_matches <- stringdist_left_join(
  rb_names_realpff_clean,
  rb_names_pff,
  by = c("player" = "pfr_player_name"),
  method = "jw",
  max_dist = 0.15,
  distance_col = "dist"
) %>%
  filter(season.x < season.y) %>%
  rename(season_college = season.x, season_draft = season.y)

# Step 2: Keep lowest-distance match(es) per draft name
best_matches <- fuzzy_matches %>%
  group_by(pfr_player_name) %>%
  filter(dist == min(dist, na.rm = TRUE)) %>%
  ungroup()


pff_cleaned <- rushing_data %>%
  filter(player %in% unique(best_matches$player)) %>% 
  select(player:attempts, touchdowns, rec_yards, rush_yards = yards, yco_attempt:yprr, season) %>% 
  left_join(best_matches %>% select(-dist), by = c("player" = "player", "season" = "season_college")) %>% 
  relocate(season:pfr_player_name, .before = player) %>% 
  filter(position %in% c("HB", "FB"))

#I did min player_id because I looked at the data, and the correct player always had a smaller player_id 
pff_cleaned_fixed <- pff_cleaned %>%
  group_by(player) %>%
  filter(player_id == min(player_id)) %>%
  ungroup() %>% 
  select(-c(ypa, touchdowns, position, team_name))

pff_pivoted <- pff_cleaned_fixed %>% 
  group_by(season_draft, pfr_player_name, player, player_id) %>%
  arrange(desc(season)) %>%
  slice_head(n = 2) %>%
  mutate(season_label = if_else(row_number() == 1, "yr2", "yr1")) %>%
  ungroup() %>%

  # Pivot longer on all stat columns except identifying info
  pivot_longer(
    cols = -c(season_draft, pfr_player_name, player, player_id, season_label),
    names_to = "stat",
    values_to = "value"
  ) %>%

  # Create combined stat name with _yr1/_yr2 suffix
  mutate(stat = paste0(stat, "_", season_label)) %>%

  # Pivot wider to get desired structure
  select(-season_label) %>%
  pivot_wider(names_from = stat, values_from = value)

 
```



#actually switch it up. use RAS (Relative Athletic Score) using a RAS Missing binary flag to incorporate those dudes that were scared of being measured

```{r}
#RAS Scores found on RAS Website, publically available at the time of access

RAS_scores <- read_csv("RB Draft History RAS.csv") %>% 
  select(-c(Link, Alltime)) %>% 
   mutate(
    RAS_missing = is.na(RAS),
    RAS = RAS,  # leave RAS as NA
    Name = case_when(
      Name == "Cameron Skattebo" ~ "Cam Skattebo",
      Name == "Matt Dayes" ~ "Matthew Dayes",
      Name == "Re'Mahn Davis" ~ "Ray Davis",
      TRUE ~ Name
    )
  )

RAS_fuzzy_matches <- stringdist_left_join(
  RAS_scores,
  rb_names_pff,
  by = c("Name" = "pfr_player_name", "Year" = "season"),
  method = "jw",
  max_dist = 0.15,
  distance_col = "dist"
) %>% 
  select(pfr_player_name, Year:RAS_missing) %>% 
  filter(!(is.na(pfr_player_name)))
  


```


```{r}
final_RB_data <- pff_pivoted %>% 
  left_join(RAS_fuzzy_matches %>% select(-College), by = c("season_draft" = "Year", "pfr_player_name")) %>% 
  left_join(drafted_rbs_with_stats_srs, by = c("pfr_player_name", "season_draft" = "season")) %>%
  relocate(round:college, .after = season_draft) %>% 
  select(-player) %>% 
  filter(season_draft > 2015) %>% 
  rename_with(~ str_replace_all(., "_yr1$", "_NCAA1")) %>%
  rename_with(~ str_replace_all(., "_yr2$", "_NCAA2")) %>%
  rename_with(~ str_replace_all(., "_1$", "_NFL1")) %>%
  rename_with(~ str_replace_all(., "_2$", "_NFL2")) %>%
  rename_with(~ str_replace_all(., "_3$", "_NFL3")) %>% 
  select(-c(player_id, season_NCAA1, season_NCAA2))

```


What I have:

-PFF College Stats
-NFL First 3 years Stats
-Measurements
-SRS

What Else do I need?:

- seperate college stats by last 2 years (this means probably cutting out 2015 draft, as they would only have 1 yr of ncaa stats)
- join into 1 df
-detail and define plan for model. what am I predicting? whats the plan after the model? 
-run model




updated plan. due to GPT reccommendations (that I should know given I just took SAL313...) I will be doing a random forest instead


```{r}
#RANDOM FOREST MODELS

# --- 0. Replace NAs in feature columns with 0 ---
non_feature_cols <- c(
  "pfr_player_name", "season_draft", "round", "pick",
  "team", "college", "receiving_yards_NFL1", "receiving_yards_NFL2", "receiving_yards_NFL3",
  "rushing_yards_NFL1", "rushing_yards_NFL2", "rushing_yards_NFL3"
)

final_RB_data <- final_RB_data %>%
  mutate(across(-all_of(non_feature_cols), ~ replace_na(.x, 0)))

# --- 1. Split the data by draft season ---
train_data <- final_RB_data %>% filter(season_draft < 2023)
test_data  <- final_RB_data %>% filter(season_draft == 2024)
pred_data  <- final_RB_data %>% filter(season_draft == 2025)

# --- 2. Define NFL targets ---
target_cols <- c("receiving_yards_NFL1", "receiving_yards_NFL2", "receiving_yards_NFL3",
                 "rushing_yards_NFL1", "rushing_yards_NFL2", "rushing_yards_NFL3")

# --- 3. Define features (everything not an identifier or NFL output) ---
exclude_cols <- c(
  target_cols,
  "pfr_player_name", "player_id", "season_draft", "round", "pick",
  "team", "college")

# --- 4. Final feature set ---
feature_cols <- setdiff(names(train_data), exclude_cols)

X_train <- train_data[, feature_cols]
X_test  <- test_data[, feature_cols]
X_pred  <- pred_data[, feature_cols]

Y_train <- train_data[, target_cols]
Y_test  <- test_data[, target_cols]

# --- 5. Store predictions ---
pred_test <- data.frame(matrix(nrow = nrow(X_test), ncol = 0))
pred_2025 <- data.frame(matrix(nrow = nrow(X_pred), ncol = 0))

# --- RECEIVING MODELS ---
rf_rec1 <- randomForest(x = X_train, y = Y_train$receiving_yards_NFL1,
                        ntree = 1000, nodesize = 1, mtry = floor(ncol(X_train) * 0.85))
pred_test$receiving_yards_NFL1 <- predict(rf_rec1, X_test)
pred_2025$receiving_yards_NFL1 <- predict(rf_rec1, X_pred)

X_train_rec2 <- X_train %>% mutate(receiving_yards_NFL1 = Y_train$receiving_yards_NFL1)
X_test_rec2  <- X_test %>% mutate(receiving_yards_NFL1 = pred_test$receiving_yards_NFL1)
X_pred_rec2  <- X_pred %>% mutate(receiving_yards_NFL1 = pred_2025$receiving_yards_NFL1)

rf_rec2 <- randomForest(x = X_train_rec2, y = Y_train$receiving_yards_NFL2,
                        ntree = 1000, nodesize = 1, mtry = floor(ncol(X_train_rec2) * 0.85))
pred_test$receiving_yards_NFL2 <- predict(rf_rec2, X_test_rec2)
pred_2025$receiving_yards_NFL2 <- predict(rf_rec2, X_pred_rec2)

X_train_rec3 <- X_train_rec2 %>% mutate(receiving_yards_NFL2 = Y_train$receiving_yards_NFL2)
X_test_rec3  <- X_test_rec2 %>% mutate(receiving_yards_NFL2 = pred_test$receiving_yards_NFL2)
X_pred_rec3  <- X_pred_rec2 %>% mutate(receiving_yards_NFL2 = pred_2025$receiving_yards_NFL2)

rf_rec3 <- randomForest(x = X_train_rec3, y = Y_train$receiving_yards_NFL3,
                        ntree = 1000, nodesize = 1, mtry = floor(ncol(X_train_rec3) * 0.85))
pred_test$receiving_yards_NFL3 <- predict(rf_rec3, X_test_rec3)
pred_2025$receiving_yards_NFL3 <- predict(rf_rec3, X_pred_rec3)

# --- RUSHING MODELS ---

rf_rush1 <- randomForest(x = X_train, y = Y_train$rushing_yards_NFL1,
                         ntree = 1000, nodesize = 1, mtry = floor(ncol(X_train) * 0.85))
pred_test$rushing_yards_NFL1 <- predict(rf_rush1, X_test)
pred_2025$rushing_yards_NFL1 <- predict(rf_rush1, X_pred)

X_train_rush2 <- X_train %>% mutate(rushing_yards_NFL1 = Y_train$rushing_yards_NFL1)
X_test_rush2  <- X_test %>% mutate(rushing_yards_NFL1 = pred_test$rushing_yards_NFL1)
X_pred_rush2  <- X_pred %>% mutate(rushing_yards_NFL1 = pred_2025$rushing_yards_NFL1)

rf_rush2 <- randomForest(x = X_train_rush2, y = Y_train$rushing_yards_NFL2,
                         ntree = 1000, nodesize = 1, mtry = floor(ncol(X_train_rush2) * 0.85))
pred_test$rushing_yards_NFL2 <- predict(rf_rush2, X_test_rush2)
pred_2025$rushing_yards_NFL2 <- predict(rf_rush2, X_pred_rush2)

X_train_rush3 <- X_train_rush2 %>% mutate(rushing_yards_NFL2 = Y_train$rushing_yards_NFL2)
X_test_rush3  <- X_test_rush2 %>% mutate(rushing_yards_NFL2 = pred_test$rushing_yards_NFL2)
X_pred_rush3  <- X_pred_rush2 %>% mutate(rushing_yards_NFL2 = pred_2025$rushing_yards_NFL2)

rf_rush3 <- randomForest(x = X_train_rush3, y = Y_train$rushing_yards_NFL3,
                         ntree = 1000, nodesize = 1, mtry = floor(ncol(X_train_rush3) * 0.85))
pred_test$rushing_yards_NFL3 <- predict(rf_rush3, X_test_rush3)
pred_2025$rushing_yards_NFL3 <- predict(rf_rush3, X_pred_rush3)


```



```{r}

pred_test <- pred_test %>%
  rename_with(~ paste0("pred_", .x))

# Join player name + truth + predictions
results_test <- test_data %>%
  select(pfr_player_name) %>%
  bind_cols(Y_test, pred_test)

# Compute RMSE for each NFL stat
rmse_table <- tibble(
  metric = c("receiving_yards_NFL1", "receiving_yards_NFL2", "receiving_yards_NFL3",
             "rushing_yards_NFL1", "rushing_yards_NFL2", "rushing_yards_NFL3"),
  RMSE = map_dbl(metric, ~ rmse(results_test[[.x]], results_test[[paste0("pred_", .x)]]))
)

print(rmse_table)

# Baseline RMSE using a naive mean model
mean_model_rmse <- rmse(
  Y_test$receiving_yards_NFL1,
  rep(mean(Y_train$receiving_yards_NFL1), nrow(Y_test))
)


results_2025 <- pred_data %>% 
  select(team, pfr_player_name) %>% 
  bind_cols(pred_2025) %>% 
  mutate(total_yards_NFL1 = receiving_yards_NFL1 + rushing_yards_NFL1,
         total_yards_NFL2 = receiving_yards_NFL2 + rushing_yards_NFL2,
         total_yards_NFL3 = receiving_yards_NFL3 + rushing_yards_NFL3,
         ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))

```


```{r}

ggplot(results_test, aes(x = receiving_yards_NFL1, y = pred_receiving_yards_NFL1)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Receiving Yards NFL1: Predicted vs Actual",
       x = "Actual", y = "Predicted")

# Rushing Yards NFL1
ggplot(results_test, aes(x = rushing_yards_NFL1, y = pred_rushing_yards_NFL1)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "darkgreen") +
  labs(title = "Rushing Yards NFL1: Predicted vs Actual",
       x = "Actual", y = "Predicted")

# --- 2. Player-level error summary (absolute error) ---
player_errors <- results_test %>%
  mutate(
    abs_error_rec1 = abs(receiving_yards_NFL1 - pred_receiving_yards_NFL1),
    abs_error_rush1 = abs(rushing_yards_NFL1 - pred_rushing_yards_NFL1)
  ) %>%
  arrange(desc(abs_error_rec1 + abs_error_rush1))  # sort by total NFL1 error

# View top players by prediction error
player_errors %>%
  select(pfr_player_name,
         receiving_yards_NFL1, pred_receiving_yards_NFL1, abs_error_rec1,
         rushing_yards_NFL1, pred_rushing_yards_NFL1, abs_error_rush1) %>%
  head(10)

```



```{r}
importance_rec <- importance(rf_rec1) %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature") %>%
  arrange(desc(IncNodePurity))

importance_rec %>%
  ggplot(aes(x = reorder(feature, IncNodePurity), y = IncNodePurity)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Feature Importance – Receiving Yards NFL1",
    x = "Feature",
    y = "Increase in Node Purity"
  ) +
  theme_minimal()


importance_rush <- importance(rf_rush1) %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature") %>%
  arrange(desc(IncNodePurity))

importance_rush %>%
  ggplot(aes(x = reorder(feature, IncNodePurity), y = IncNodePurity)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(
    title = "Feature Importance – Rushing Yards NFL1",
    x = "Feature",
    y = "Increase in Node Purity"
  ) +
  theme_minimal()

```
```{r}

importance_rec <- importance(rf_rec1) %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature") %>%
  mutate(model = "Receiving Yards NFL1")

importance_rush <- importance(rf_rush1) %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature") %>%
  mutate(model = "Rushing Yards NFL1")

# Combine and select top 15 features from each
importance_combined <- bind_rows(importance_rec, importance_rush) %>%
  group_by(model) %>%
  ungroup()

importance_combined %>%
  group_by(model) %>%
  ungroup() %>%
  ggplot(aes(x = reorder_within(feature, IncNodePurity, model), y = IncNodePurity, fill = model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ model, scales = "free") +
  coord_flip() +
  scale_x_reordered() +
  labs(
    title = "Feature Importance for Receiving vs Rushing Yards NFL1",
    x = "Feature",
    y = "Increase in Node Purity"
  ) +
  theme_minimal()



```

